Bow shocks, bow waves, and dust waves

### 2019-01-15 papers/Papers1234/sec-dust-bow-intro.tex
Revise intro and promote subsections to sections

Curved emission arcs around stars \citep[e.g.,][]{Gull:1979a} are
often interpreted as \textit{bow shocks}, due to a supersonic
hydrodynamic interaction between the star's wind and an external
stream. This stream may be due to the star's own motion or to an
independent flow, such as an \hii{} region in the champagne phase
\citep{Tenorio-Tagle:1979a}, or another star's wind
\citep{Canto:1996}. However, an alternative interpretation in some
cases may be a radiation-pressure driven bow wave, as first proposed
by \citet[\S\textsc{vi}]{van-Buren:1988a}.  In this scenario, photons
emitted by the star are absorbed by dust grains in the incoming
stream, with the resultant momentum transfer being sufficient to
decelerate and deflect the grains within a certain distance from the
star, forming a dust-free, bow-shaped cavity with an enhanced dust
density at its edge.  Two regimes are possible, depending on the
strength of coupling between the gas (or plasma) and the dust.  In the
strong-coupling regime, gas--grain drag decelerates the gas along with
the dust, forming a shocked gas shell in a similar fashion to the
wind-driven bow shock case.  In the weak-coupling regime, the gas
stream is relatively unaffected and the dust temporarily decouples to
form a dust-only shell.  This second case has recently been studied in
detail in the context of the interaction of late O-type stars (which
have only weak stellar winds) with dusty photoevaporation flows inside
\hii{} regions \citep{Ochsendorf:2014a, Ochsendorf:2014b,
  Ochsendorf:2015a}.  We follow the nomenclature proposed by
\citet{Ochsendorf:2014b}, in which \textit{dust wave} refers to the
weak coupling case and \textit{bow wave} to the strong coupling case.
More complex, hybrid scenarios are also possible, such as that studied
by \citet{van-Marle:2011a}, where a hydrodynamic bow shock forms, but
the larger dust grains that accompany the stellar wind pass right
through the shocked gas shell, and form their own dust wave at a
larger radius.

@@

In \citet[][hereafter \PaperI{}]{Tarango-Yong:2018a}, we proposed a
new two-dimensional classification scheme for bow shapes: the
projected planitude--alatude, or \(\Pi'\)--\(\Lambda'\), diagram.  Planitude
measures the flatness of the bow's apex, while alatude measures the
openness of the bow's wings.  Both are dimensionless ratios of lengths
that can be estimated from observational images.  We have analyzed the
inclination-dependent tracks on the \(\Pi'\)--\(\Lambda'\) plane for simple
geometric shapes (spheroids, paraboloids, hyperboloids) and for
thin-shell hydrodynamic bow shock models (wilkinoid, cantoids,
ancantoids).  In this paper, we will do the same for simple models of
radiation-driven dust waves (dragoids) and bow waves (trapoids).

The paper is organized as follows.
%
In \S~\ref{sec:shape-dust-wave} we do the same for simple models of a
dusty radiation bow wave (dragoids), including the effects of
gas-grain drag.
%
In \S~\ref{sec:perturbed-bows} we investigate the effects on the
planitude--alatude plane of small-amplitude perturbations to the bow
shape.
%

\section{Different types of bow}
\label{sec:different-types-bow}

@@

In this section, we investigate the different types of bow interaction
that will occur in different regions of parameter space. We will
mainly treat the canonical case\footnote{%
  Variant cases with differing arrangements of dust and radiation
  sources are treated in \S~\ref{sec:case-inside-out}.} %
of a bow around a star of bolometric luminosity, \(L\), with a
radiatively driven wind, which is immersed in an external stream of
gas and dust with density, \(\rho\), and velocity, \(v\).  The size and
shape of the bow is determined by a generalized balance of pressure
(or, equivalently, momentum) between internal and external sources.
We assume that the stream is supersonic and super-alfvenic, so that
the external pressure is dominated by the ram pressure: \(\rho v^2\).

\subsection{Strong gas-grain coupling}
\label{sec:strong-gas-grain}

We first consider the case where the dust grains and gas are perfectly
coupled by collisions.\footnote{%
  Cases where this assumption does not hold are investigated below in
  \S~\ref{sec:imperf-coupl-betw}.} %
Although dust grains typically constitute only a small fraction
\(Z\grain \sim 0.01\) of the mass of the external stream, they
nevertheless dominate the broad-band opacity at FUV, optical and IR
wavelengths if they are present.\footnote{%
  At EUV wavelengths (\(\lambda < \SI{912}{\angstrom}\)), gas opacity
  dominates if the hydrogen neutral fraction is larger than
  \(\approx 0.001\), see discussion of ionization front trapping
  below.} %
The strong coupling assumption means that all the radiative forces
applied to the dust grains are directly felt by the gas also.

\subsubsection{Bows supported by radiation and wind}
\label{sec:three-bow-regimes}

@@

\subsubsection{Effects of stellar gravity}
\label{sec:effects-gravity}

@@

\subsubsection{Ionization state of the bow shell}
\label{sec:trapp-ioniz-front}

@@

\subsubsection{Radiative cooling lengths}
\label{sec:radi-cool-lengths}
\newcommand\M{\ensuremath{\mathcal{M}}}
In this section, we calculate whether the radiative cooling is
sufficiently rapid behind the bow shock to allow the formation of a
thin, dense shell.  Since cooling is least efficient at low densities,
we will assume that the wind bow shock regime applies unless otherwise
specified. We label quantities just outside the shock by the subscript
``0'', quantities just inside the shock (after thermalization, but
before any radiative cooling) by the subscript ``1'', and quantities
after the gas has cooled back to the photoionization equilibrium
temperature by the subscript ``2''.  Assuming a ratio of specific
heats, \(\gamma = 5/3\), the relation between the pre-shock and immediate
post-shock quantities is
\begin{align}
  % \M_1 &= \left(\frac{\M_0^2 + 3} {5\M_0^2 - 1}\right)^{1/2} \\
  \label{eq:shock-n-jump}
  \frac{n_1}{n_0} &= \frac{4 \M_0^2} {\M_0^2 + 3} \\
  \label{eq:shock-T-jump}
  \frac{T_1}{T_0} &= \tfrac1{16} \bigl( 5\M_0^2 - 1 \bigr) \bigl( 1 + 3/\M_0^2 \bigr) \\
  \label{eq:shock-v-jump}
  \frac{v_1}{v_0} &= \left(\frac{n_1}{n_0}\right)^{-1} \ ,
\end{align}
where \(\M_0 = v_0 / \sound\).  The cooling length of the post-shock
gas can be written as
\newcommand\cool{\ensuremath{_{\text{cool}}}}
\begin{equation}
  \label{eq:dcool}
  d\cool = \frac{3 P_1 v_1} { 2 \bigl(  \mathcal{L}_1 - \mathcal{G}_1 \bigr) }\ ,  
\end{equation}
where \(P_1\) is the thermal pressure and \(\mathcal{L}_1\),
\(\mathcal{G}_1\) are the volumetric radiative cooling and heating
rates.  For fully photoionized gas, we have
\(P_1 \approx 2 n_1 k T_1\), \(\mathcal{L}_1 = n_1^2 \Lambda(T_1)\), and
\(\mathcal{G}_1 = n_1^2 \Gamma(T_1)\), where \(\Lambda(T)\) is the cooling
coefficient, which is dominated by metal emission lines that are
excited by electron collisions, and \(\Gamma(T)\) is the heating
coefficient, which is dominated by hydrogen photo-electrons
\citep{Osterbrock:2006a}. The cooling coefficient has a maximum around
\SI{e5}{K}, and for typical ISM abundances can be approximated as
follows:
\begin{align}
  \label{eq:cooling-coefficient}
  \Lambda_{\text{warm}} &= \num{3.3e-24} \, T_4^{2.3} \, \si{erg.cm^{-3}.s^{-1}}\\
  \Lambda_{\text{hot}} &= \num{e-20} \, T_4^{-1}\, \si{erg.cm^{-3}.s^{-1}} \\
  \Lambda &= \left( \Lambda_{\text{warm}}^{-k} +  \Lambda_{\text{hot}}^{-k} \right)^{-1/k}
      \quad \text{with} \quad k = 3 \ ,
\end{align}
which is valid in the range \(0.7 < T_4 < 1000\).  We approximate the heating coefficient as
\begin{equation}
  \label{eq:heating-coefficient}
  \Gamma = \num{1.77e-24} \, T_4^{-1/2} \, \si{erg.cm^{-3}.s^{-1}} \ ,
\end{equation}
where the coefficient is chosen so as to give \(\Gamma = \Lambda\) at
an equilibrium temperature of \(T_4 = 0.8\).

@@

\subsection{Imperfect coupling between gas and dust}
\label{sec:imperf-coupl-betw}

@@

\subsubsection{Drag force on grains}
\label{sec:drag-force-grains}

@@

\subsubsection{Gas--grain separation: drift and rip}
\label{sec:gas-grain-separ}

@@

\subsubsection{Existence conditions for dust waves}
\label{sec:exist-cond-separ}

@@

\subsubsection{Grain trajectories along the symmetry axis}
\label{sec:grain-traj-along}

@@

\subsubsection{Back reaction on the gas flow}
\label{sec:back-reaction-gas}

@@

\subsubsection{Magnetic effects on grain trajectories }
\label{sec:magn-effects-grain}

>>>>

Curved emission arcs around stars \citep[e.g.,][]{Gull:1979a} are
often interpreted as \textit{bow shocks}, due to a supersonic
hydrodynamic interaction between the star's wind and an external
stream. This stream may be due to the star's own motion or to an
independent flow, such as an \hii{} region in the champagne phase
\citep{Tenorio-Tagle:1979a}, or another star's wind
\citep{Canto:1996}. However, an alternative interpretation in some
cases may be a radiation-pressure driven bow wave, as first proposed
by \citet[\S\textsc{vi}]{van-Buren:1988a}.  In this scenario, photons
emitted by the star are absorbed by dust grains in the incoming
stream, with the resultant momentum transfer being sufficient to
decelerate and deflect the grains within a certain distance from the
star, forming a dust-free, bow-shaped cavity with an enhanced dust
density at its edge.

Two regimes are possible, depending on the strength of coupling
between the gas (or plasma) and the dust.  In the strong-coupling
regime, gas--grain drag decelerates the gas along with the dust.  If
the stream is optically thin to the star's ultraviolet radiation, then
the deceleration occurs gradually over a range of radii, forming a
relatively thick shell.  On the other hand, if the stream is optically
thick, then a shocked gas shell forms in a similar fashion to the
wind-driven bow shock case, except internally supported by trapped
radiation instead of shocked stellar wind.  In the weak-coupling
regime, the gas stream is relatively unaffected and the dust
temporarily decouples to form a dust-only shell.  This second case has
recently been studied in detail in the context of the interaction of
late O-type stars (some of which have very weak stellar winds) with
dusty photoevaporation flows inside \hii{} regions
\citep{Ochsendorf:2014a, Ochsendorf:2014b, Ochsendorf:2015a}.  We
follow the nomenclature proposed by \citet{Ochsendorf:2014b}, in which
\textit{dust wave} refers to the weak coupling case and \textit{bow
  wave} to the strong coupling case.  More complex, hybrid scenarios
are also possible, such as that studied by \citet{van-Marle:2011a},
where a hydrodynamic bow shock forms, but the larger dust grains that
accompany the stellar wind pass right through the shocked gas shell,
and form their own dust wave at a larger radius.

@@

% In \citet[][hereafter \PaperI{}]{Tarango-Yong:2018a}, we proposed a
% new two-dimensional classification scheme for bow shapes: the
% projected planitude--alatude, or \(\Pi'\)--\(\Lambda'\), diagram.  Planitude
% measures the flatness of the bow's apex, while alatude measures the
% openness of the bow's wings.  Both are dimensionless ratios of lengths
% that can be estimated from observational images.  We have analyzed the
% inclination-dependent tracks on the \(\Pi'\)--\(\Lambda'\) plane for simple
% geometric shapes (spheroids, paraboloids, hyperboloids) and for
% thin-shell hydrodynamic bow shock models (wilkinoid, cantoids,
% ancantoids).  In this paper, we will do the same for simple models of
% radiation-driven dust waves (dragoids) and bow waves (trapoids).

In this paper, we develop simple physical models to show in detail
when and how these different interaction regimes apply when varying
the parameters of the star, the dust grains, and the ambient stream.
We concentrate primarily on the case of luminous early type stars,
where dust is present only in the ambient stream, and not in the
stellar wind.
%
The paper is organized as follows.
%
In \S~\ref{sec:strong-gas-grain} we consider the case where the grains
are perfectly coupled to the gas via collisions, investigating the
relative importance of wind and radiation in providing internal
support for the bow shell as a function of the density and velocity of
the ambient stream, as well as calculating the shell's ionization
state and radiative cooling efficiency.
%
In \S~\ref{sec:imperf-coupl-betw} we calculate in detail the grain
charging and dynamics, subject to radiation and drag forces, in order
to determine the rip point where gas--grain coupling catastrophically
breaks down.  This is used to determine existence conditions for the
presence of a decoupled dust wave.
%
In \S~\ref{sec:magn-effects-grain} we consider the additional coupling
between the grains and the plasma's magnetic field, and how this
effects the existence and structure of dust waves.
%
In \S~\ref{sec:summary-discussion} we briefly discuss the application
of our models to observed bows.
%
In \S~\ref{sec:conclusions} we summarise the conclusions of our study. 
%
A series of appendices provide additional technical information on
dust physics (Appendix~\ref{sec:cloudy-models-dust}), analytical
models of the shape of dust waves in the drag-free limit
(Appendix~\ref{sec:gas-free-bow}), and the numerical calculation of
grain trajectories (Appendix~\ref{sec:equat-moti-grains}).

% In \S~\ref{sec:shape-dust-wave} we do the same for simple models of a
% dusty radiation bow wave (dragoids), including the effects of
% gas-grain drag.
% %
% In \S~\ref{sec:perturbed-bows} we investigate the effects on the
% planitude--alatude plane of small-amplitude perturbations to the bow
% shape.
% %

@@

\section{Strong gas-grain coupling}
\label{sec:strong-gas-grain}

We consider the canonical case of a bow around a star of bolometric
luminosity, \(L\), with a radiatively driven wind, which is immersed
in an external stream of gas and dust with density, \(\rho\), and
velocity, \(v\).  The size and shape of the bow is determined by a
generalized balance of pressure (or, equivalently, momentum) between
internal and external sources.  We assume that the stream is
supersonic and super-alfvenic, so that the external pressure is
dominated by the ram pressure: \(\rho v^2\).  We first consider the case
where the dust grains and gas are perfectly coupled by collisions.

Although dust grains typically constitute only a small fraction
\(Z\grain \sim 0.01\) of the mass of the external stream, they
nevertheless dominate the broad-band opacity at FUV, optical and IR
wavelengths if they are present.\footnote{%
  At EUV wavelengths (\(\lambda < \SI{912}{\angstrom}\)), gas opacity
  dominates if the hydrogen neutral fraction is larger than
  \(\approx 0.001\), see discussion of ionization front trapping
  below.} %
The strong coupling assumption means that all the radiative forces
applied to the dust grains are directly felt by the gas also.

\subsection{Bows supported by radiation and wind}
\label{sec:three-bow-regimes}

@@

\subsection{Effects of stellar gravity}
\label{sec:effects-gravity}

@@

\subsection{Ionization state of the bow shell}
\label{sec:trapp-ioniz-front}

@@

\subsection{Radiative cooling lengths}
\label{sec:radi-cool-lengths}
\newcommand\M{\ensuremath{\mathcal{M}}}
%
In this section, we calculate whether the radiative cooling is
sufficiently rapid behind the bow shock to allow the formation of a
thin, dense shell.  Since this is mainly a concern at low densities,
where cooling is least efficient, we will assume that the wind bow
shock regime applies unless otherwise specified. We label quantities
just outside the shock by the subscript ``0'', quantities just inside
the shock (after thermalization, but before any radiative cooling) by
the subscript ``1'', and quantities after the gas has cooled back to
the photoionization equilibrium temperature by the subscript ``2''.
Assuming a ratio of specific heats, \(\gamma = 5/3\), the relation between
the pre-shock and immediate post-shock quantities is
\begin{align}
  % \M_1 &= \left(\frac{\M_0^2 + 3} {5\M_0^2 - 1}\right)^{1/2} \\
  \label{eq:shock-n-jump}
  \frac{n_1}{n_0} &= \frac{4 \M_0^2} {\M_0^2 + 3} \\
  \label{eq:shock-T-jump}
  \frac{T_1}{T_0} &= \tfrac1{16} \bigl( 5\M_0^2 - 1 \bigr) \bigl( 1 + 3/\M_0^2 \bigr) \\
  \label{eq:shock-v-jump}
  \frac{v_1}{v_0} &= \left(\frac{n_1}{n_0}\right)^{-1} \ ,
\end{align}
where \(\M_0 = v_0 / \sound\).  The cooling length of the post-shock
gas can be written as
\newcommand\cool{\ensuremath{_{\text{cool}}}}
\begin{equation}
  \label{eq:dcool}
  d\cool = \frac{3 P_1 v_1} { 2 \bigl(  \mathcal{L}_1 - \mathcal{G}_1 \bigr) }\ ,  
\end{equation}
where \(P_1\) is the thermal pressure and \(\mathcal{L}_1\),
\(\mathcal{G}_1\) are the volumetric radiative cooling and heating
rates.  For fully photoionized gas, we have
\(P_1 \approx 2 n_1 k T_1\), \(\mathcal{L}_1 = n_1^2 \Lambda(T_1)\), and
\(\mathcal{G}_1 = n_1^2 \Gamma(T_1)\), where \(\Lambda(T)\) is the cooling
coefficient, which is dominated by metal emission lines that are
excited by electron collisions, and \(\Gamma(T)\) is the heating
coefficient, which is dominated by hydrogen photo-electrons
\citep{Osterbrock:2006a}. The cooling coefficient has a maximum around
\SI{e5}{K}, and for typical ISM abundances can be approximated as
follows:
\begin{align}
  \label{eq:cooling-coefficient}
  \Lambda_{\text{warm}} &= \num{3.3e-24} \, T_4^{2.3} \, \si{erg.cm^{-3}.s^{-1}}\\
  \Lambda_{\text{hot}} &= \num{e-20} \, T_4^{-1}\, \si{erg.cm^{-3}.s^{-1}} \\
  \Lambda &= \left( \Lambda_{\text{warm}}^{-k} +  \Lambda_{\text{hot}}^{-k} \right)^{-1/k}
      \quad \text{with} \quad k = 3 \ ,
\end{align}
which is valid in the range \(0.7 < T_4 < 1000\).  We approximate the heating coefficient as
\begin{equation}
  \label{eq:heating-coefficient}
  \Gamma = \num{1.77e-24} \, T_4^{-1/2} \, \si{erg.cm^{-3}.s^{-1}} \ ,
\end{equation}
where the coefficient is chosen so as to give \(\Gamma = \Lambda\) at
an equilibrium temperature of \(T_4 = 0.8\).

@@

\section{Imperfect coupling between gas and dust}
\label{sec:imperf-coupl-betw}

@@

\subsection{Drag force on grains}
\label{sec:drag-force-grains}

@@

\subsection{Gas--grain separation: drift and rip}
\label{sec:gas-grain-separ}

@@

\subsection{Existence conditions for dust waves}
\label{sec:exist-cond-separ}

@@

\subsection{Grain trajectories along the symmetry axis}
\label{sec:grain-traj-along}

@@

\subsection{Back reaction on the gas flow}
\label{sec:back-reaction-gas}

@@

\section{Magnetic effects on grain trajectories }
\label{sec:magn-effects-grain}
